---
title: "Getting started with summarisebig"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with summarisebig}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

```{r setup}
library(arrow)
library(dplyr)
library(summarisebig)
```

# What problem does `summarisebig` solve?

Arrow is very good at filtering, transforming, grouping, and aggregating data
without first loading an entire dataset into an R data frame. The difficult case
is when a grouped calculation eventually needs ordinary R code that Arrow cannot
execute directly.

`summarise_big()` uses the following hierarchy:

1. Try to compute the complete grouped summary in Arrow.
2. If the answer can be reconstructed from compact Arrow-computable state, use
   `.strategy = "map_reduce"` and let R perform only the final combination.
3. If the custom R function genuinely needs raw observations, materialize
   complete groups in bounded chunks with `parallel_chunks` or `shared_chunk`.

The examples below use a tiny Parquet dataset so that they are safe to run in a
vignette. The same interface is intended for datasets much larger than memory.

```{r data}
d <- data.frame(
  grp = rep(c("A", "B", "C"), each = 4),
  row_id = rep(1:4, 3),
  x = c(1, 2, 4, 8, 10, 12, 14, 18, 3, 6, 9, 15),
  y = c(2, 3, 5, 9, 9, 13, 15, 20, 5, 5, 11, 14),
  txt = c("alpha", "bravo", "charlie", "delta",
          "echo", "foxtrot", "golf", "hotel",
          "india", "juliet", "kilo", "lima")
)

path <- tempfile("summarisebig-vignette-")
arrow::write_dataset(d, path, format = "parquet")
ds <- arrow::open_dataset(path)
```

# 1. Let Arrow do everything when it can

The ordinary strategies first try the entire expression in Arrow. For a
supported grouped summary, raw observations never need to become an R data
frame.

```{r arrow-fast-path}
summarise_big(
  ds,
  result = mean(x),
  .by = grp
) |>
  arrange(grp)
```

A fairly complicated expression can still remain entirely in Arrow if all its
parts can be translated by Arrow. Complexity by itself is therefore not a
reason to use MapReduce or materialization.

# 2. Go as far as possible in Arrow, then finalize in R

Suppose a final result can be reconstructed from a few group-level quantities.
For sample variance we only need the number of observations, the sum, and the
sum of squares. Arrow calculates those quantities on the full dataset and R
receives only the compact state table.

```{r map-reduce}
summarise_big(
  ds,
  .by = grp,
  .strategy = "map_reduce",
  .map_reduce = list(
    n = ~ dplyr::n(),
    sx = ~ sum(x),
    sx2 = ~ sum(x * x)
  ),
  .finalize = function(state) {
    state |>
      mutate(
        variance = (sx2 - sx * sx / n) / (n - 1)
      ) |>
      select(grp, variance)
  }
) |>
  arrange(grp)
```

`map_reduce` does **not** discover the decomposition automatically. The user
supplies the Arrow-computable building blocks and the R finalizer. This is often
the most attractive route after pure Arrow because the large-data work remains
inside Arrow and only a few numbers per group cross into R.

# 3. Arbitrary R functions: `parallel_chunks`

Some summaries genuinely need the raw values. Here is an intentionally ordinary
R function based on exact quantiles:

```{r custom-function}
interquartile_span <- function(x) {
  q <- stats::quantile(
    x,
    probs = c(0.25, 0.75),
    names = FALSE,
    type = 7
  )
  q[[2]] - q[[1]]
}
```

We disable the Arrow fast path in this example so that the materialized route is
exercised explicitly. Complete groups are packed into bounded chunks and
separate chunks can be processed by different Mirai workers.

```{r parallel-chunks}
summarise_big(
  ds,
  result = interquartile_span(x),
  .by = grp,
  .strategy = "parallel_chunks",
  .workers = 2,
  .chunk_rows = 4,
  .try_arrow = FALSE
) |>
  arrange(grp)
```

For expensive user-defined R functions this strategy can benefit substantially
from multiple workers. For very cheap functions, parallel startup and scheduling
overhead can dominate, so more workers are not automatically faster.

# 4. Lower-memory materialization: `shared_chunk`

`shared_chunk` materializes only one disk chunk at a time. It first sorts the
chunk so that each group is contiguous, then divides complete groups into
row-balanced tasks. With multiple workers and the optional `mori` package, the
chunk can be shared rather than copied to every worker.

The following order-sensitive string summary also demonstrates why `.order_by`
exists: Arrow Dataset physical row order should not be treated as meaningful.

```{r shared-chunk}
initials_in_order <- function(txt) {
  paste0(substr(txt, 1, 1), collapse = "")
}

summarise_big(
  ds,
  result = initials_in_order(txt),
  .by = grp,
  .order_by = "row_id",
  .strategy = "shared_chunk",
  .workers = 1,
  .chunk_rows = 8,
  .task_rows = 4,
  .try_arrow = FALSE
) |>
  arrange(grp)
```

Using one worker above keeps the vignette independent of optional shared-memory
support. With `mori` installed, the same call can use multiple workers:

```{r shared-chunk-parallel, eval=FALSE}
summarise_big(
  ds,
  result = initials_in_order(txt),
  .by = grp,
  .order_by = "row_id",
  .strategy = "shared_chunk",
  .workers = 2,
  .chunk_rows = 8,
  .task_rows = 4,
  .try_arrow = FALSE
)
```

# Choosing a strategy

A useful decision rule is:

| Situation | Preferred route |
|---|---|
| Arrow can translate the complete expression | Arrow fast path |
| Final answer needs R but compact Arrow state is sufficient | `map_reduce` |
| Arbitrary R needs raw rows and throughput matters most | `parallel_chunks` |
| Arbitrary R needs raw rows and peak memory matters most | `shared_chunk` |

For the two materialized strategies, a complete statistical group is never split.
Therefore a genuinely opaque R calculation still requires each individual group
to fit in memory. If one group is itself too large, the calculation must be
reformulated in Arrow or as a compact reduction if that is mathematically
possible.

```{r cleanup, include=FALSE}
unlink(path, recursive = TRUE, force = TRUE)
```
